NumPy---Array Computation for Python

### 2019-12-21 body.tex
Polishing intro sections

%NumPy is the fundamental array computation library for the Python
%ecosystem \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
Since its first stable release in October 2006, NumPy has unified
the scientific Python community and now underpins almost every library
that does numerical computation, including SciPy\cite{virtanen2019scipy},
pandas\cite{mckinney-proc-scipy-2010}, scikit-learn\cite{pedregosa2011scikit},
and scikit-image\cite{vanderwalt2014scikit}.
Not surprisingly, many important scientific research projects and
results crucially rely on NumPy.
It plays an essential role in many research analysis pipelines in
fields as diverse as physics, biology, astronomy, neuroscience,
material science, engineering, and chemistry.
% w/ citations
In particular, NumPy played a fundamental role in discovering
gravitational waves\cite{pycbc}, imaging black
holes\cite{eht-imaging}, and exploring the solar system \cite{jenness2018lsst}.

@@

NumPy consists of the \code{ndarray}-object along with utility functions that
operate on it.
Because of its inherent simplicity—being a pointer to memory with some
associated meta-data about shape, data-type, and so forth—the NumPy array is
the {\it de facto} exchange format for array data in Python.
NumPy, having been wildly successful as a library and array exchange
project, now also has become a standard language for expressing array and
tensor computations.
It has such widespread adoption that not only the array object but also its
{\it Application Programming Interface} (API) has become ubiquitous, and is
implemented by, among others, popular tensor computation libraries\cite{pytorch}.

@@

Before NumPy, two different Python array packages—Numeric and NumArray—existed.
The first began in the mid-90s and was developed by Jim Fulton, Jim Hugunin,
Paul Dubois, Konrad Hinsen, Travis Oliphant, and others, and was aimed at fast
manipulation of many small arrays.
By the beginning of the millennium, several libraries built on top of Numeric 
the provided fundamental scientific routines were combined to form SciPy.
Due to the poor performance of Numeric on large data sets, NumArray was
developed at the Space Telescope Science Institute (STScI) to handle
large astronomical images such as those produced by the Hubble Space Telescope.
It was initially considered the successor to Numeric.
However, it performed poorly on small arrays.
This divide posed a challenge for SciPy as it wasn't clear how to support
both libraries.
Working with the NumArray and SciPy community, the maintainer of Numeric,
Travis Oliphant, started a major rewrite that aimed to be a ``best of both
worlds'' unification of the two projects.  %This became known as NumPy.
This merger of Numeric and NumArray was called NumPy.

The creation of NumPy spurred renewed development in both SciPy and the larger
scientific Python ecosystem and heralded the current era of wide-spread use of
Python for scientific computing.
To coordinate work as a global community, to ensure accuracy of computed
results, and to allow modification without breakage, the NumPy (and SciPy) community
pioneered best software engineering practices in the scientific Python
community  \cite{millman2014developing}.
For example, NumPy early on adopted distributed revision control, unit testing,
continuous integration, code review, and standardized executable documentation.
It also developed a standard for ``docstrings''—text describing each NumPy function.
This quickly became the standard for the larger scientific Python community.

As the scientific Python community grew and new challenges arose, Travis Oliphant
eventually stepped back from leading NumPy to found a company and start
new projects including Numba---an accelerated Python compiler that could vastly
accelerate NumPy expressions.
Development of NumPy continued by a core group that informally operated by consensus.

@@

For many years, NumPy had no dedicated funding, development
time was mostly contributed freely by students and researchers in
their spare time.  In-person meetings were sponsored off of
individual's research grants, and while industry sponsored
contributions were made, notably those by Mark Wiebe when he was
sponsored by Travis Oliphant at Continuum Analytics, mostly the
package had rapid adoption and became central to scientific computing
without external investment.

@@

% CZI -- NumFOCUS/QuanSight
% https://chanzuckerberg.com/eoss/proposals/strengthening-numpys-foundations-growing-beyond-code/
% NumPy and OpenBLAS received $195,000 in 2019
% https://labs.quansight.org/blog/2019/11/numpy-openblas-CZI-grant/
NumPy received another grant for \$195,000 from the Chan Zuckerberg
Initiative at the end of 2019.
Ralf Gommers is the PI.
This grant, building on the technical work supported by the previous
grant, will focus on better serving NumPy's large number of beginner
to intermediate level users and on growing the community of NumPy
contributors.
It also will provide some support for OpenBLAS (on which NumPy depends for
accelerated linear algebra).

The project also receives around 10K USD per year from Tidelift, which is
used to fund documentation and website developers.

@@

The weekly community calls alternate weekly between triage meetings and
higher level discussion.  The calls not only involve developers from
the community, but provide a venue for vendors and other external
groups to provide input.  For example, after Intel produced a forked
version of NumPy, one of their developers joined a call to discuss
community concerns.

@@

NumPy is currently maintained by a group of 23 contributors with commit rights
to the NumPy code base. Out of these 17 maintainers were active in 2019 and 4
of those were paid to work on the project full-time.
Additionally, there are a few long term developers who contributed and maintain
specific parts of NumPy, but are not officially maintainers.
At a release cycle of about every half year, the five recent releases in the years
2018 and 2019 have averaged about 450~pull requests~(PRs) each.\footnote{
    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
    All data provided here is based on the development which happened through github
    pull requests. In some cases contributions by maintainers may not be categorized as such.}
% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
With each release attracting more than a hundred new contributors.

@@

>>>>

%NumPy is the fundamental array computation library for the Python
%ecosystem \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
With its first stable release in October 2006, NumPy unified
the scientific Python community and now underpins almost every library
that does numerical computation, including SciPy\cite{virtanen2019scipy},
pandas\cite{mckinney-proc-scipy-2010}, scikit-learn\cite{pedregosa2011scikit},
and scikit-image\cite{vanderwalt2014scikit}.
Not surprisingly, many important scientific research projects and
results crucially rely on NumPy.
It plays an essential role in many research analysis pipelines in
fields as diverse as physics, biology, astronomy, neuroscience,
material science, engineering, and chemistry.
% w/ citations
In particular, NumPy played a fundamental role in discovering
gravitational waves\cite{pycbc} and the first imaging of a black
hole\cite{eht-imaging}, and lets us explore our
universe\cite{jenness2018lsst}.

@@

NumPy consists of the \code{ndarray}-object along with utility functions that
operate on it.
Because of its inherent simplicity—being a pointer to memory with some
associated meta-data about shape, data-type, and so forth—the NumPy array is
the {\it de facto} exchange format for array data in Python.
The library has such widespread adoption that not only the array object but also its
{\it Application Programming Interface} (API) has become ubiquitous as
a language for tensor computation---witnessed by its use in popular
deep learning libraries such as PyTorch\cite{pytorch}.

@@

Before NumPy, two different Python array packages—Numeric and NumArray—existed.
The first began in the mid-90s and was developed by Jim Fulton, Jim Hugunin,
Paul Dubois, Konrad Hinsen, Travis Oliphant, and others, and was aimed at fast
manipulation of many small arrays.
By the beginning of the millennium, several libraries built on top of Numeric 
to provide fundamental scientific routines were combined into SciPy.
Due to the poor performance of Numeric on large data sets, NumArray was
developed at the Space Telescope Science Institute (STScI) to handle
large astronomical images such as those produced by the Hubble Space Telescope.
However, while initially considered the successor to Numeric, NumArray was no longer
performant on multiple small arrays.
This divide posed a challenge for SciPy as it wasn't clear how to support
both libraries.
Working with the NumArray and SciPy communities, the then maintainer of Numeric,
Travis Oliphant, embarked upon a major rewrite that aimed to be a ``best of both
worlds'' unification of Numeric and NumArray. This merger resulted in NumPy.

The creation of NumPy spurred renewed development in both SciPy and the larger
scientific Python ecosystem and heralded the current era of wide-spread use of
Python for scientific computing.
To coordinate work as a global community, to ensure accuracy of computed
results, and to allow modification without breakage, the NumPy (and SciPy) teams
pioneered best software engineering practices in the scientific Python
community\cite{millman2014developing}.
For example, NumPy early on adopted distributed revision control, unit testing,
continuous integration, code review, and executable documentation.
It also developed a standard for ``docstrings''—text describing each NumPy function.
This quickly became a documentation standard for the larger scientific Python community.

Travis Oliphant eventually stepped back from leading NumPy to found a
company and to start new projects including Numba---an accelerated Python
compiler that could vastly accelerate NumPy expressions.  Development
of NumPy continued by a core group that informally operated by
consensus.

% While the scientific Python community grew and new challenges arose,

@@

For many years, NumPy had no dedicated funding, and development time
was mostly contributed freely by students and researchers in their
spare time.  In-person meetings were sponsored off of individual's
research grants, and while industry sponsored contributions were made,
notably those by Mark Wiebe when he was sponsored by Travis Oliphant
at Continuum Analytics, the package's rapid adoption in scientific
computing took place without external investment.

@@

% CZI -- NumFOCUS/QuanSight
% https://chanzuckerberg.com/eoss/proposals/strengthening-numpys-foundations-growing-beyond-code/
% NumPy and OpenBLAS received $195,000 in 2019
% https://labs.quansight.org/blog/2019/11/numpy-openblas-CZI-grant/
NumPy received another grant for \$195,000 from the Chan Zuckerberg
Initiative at the end of 2019 with Ralf Gommers as the PI.
This grant, building on the technical work supported by the previous
grant, will focus on better serving NumPy's large number of beginning
to intermediate level users and on growing the community of NumPy
contributors.
It will also provide support to OpenBLAS, on which NumPy depends for
accelerated linear algebra.

Finally, the project receives around 10K USD annually from Tidelift, which is
typically used to fund documentation writers and website developers.

@@

Weekly community calls alternate between triage and
higher level discussion.  The calls not only involve developers from
the community, but provide a venue for vendors and other external
groups to provide input.  For example, after Intel produced a forked
version of NumPy, one of their developers joined a call to discuss
community concerns.

@@

NumPy is currently maintained by a group of 23 contributors with commit rights
to the NumPy code base. Out of these, 17 maintainers were active in
2019, 4 of whom were paid to work on the project full-time.
Additionally, there are a few long term developers who contributed and maintain
specific parts of NumPy, but are not officially maintainers.
At a release cycle of about every half year, the five recent releases in the years
2018 and 2019 have averaged about 450~pull requests~(PRs) each,\footnote{
    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
    All data provided here is based on the development which happened through GitHub
    pull requests. In some cases contributions by maintainers may not be categorized as such.}
% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
with each release attracting more than a hundred new contributors.

@@

% -- The paragraphs above and below can be tightened up a bit still --
